---
title: "Exam_4"
output: html_document
---

#### Exam 4 Biol 490

#### Redo of Exam 3
```{r}
library(dada2)
library(vegan)
library(tidyr)
```

### Part I

## This is my bash code showing that I converted fastq to fasta and counted the lines:
```{bash eval=FALSE, include=TRUE}
STUC02MMPMLFH00:Exam_3 jamescutler$ ls -ahl
total 21536
drwxr-xr-x  8 jamescutler  staff   256B Apr 26 21:24 .
drwxr-xr-x  5 jamescutler  staff   160B Apr 24 10:15 ..
-rw-r--r--  1 jamescutler  staff   3.4K Apr 24 10:15 Exam_3.Rmd
-rw-r--r--  1 jamescutler  staff   723K Apr 24 10:15 Exam_3.html
-rw-r--r--  1 jamescutler  staff   4.1M Apr 24 10:15 Sample1.fastq
-rw-r--r--  1 jamescutler  staff   4.9M Apr 24 10:15 Sample2.fastq
-rw-r--r--  1 jamescutler  staff   3.0K Apr 24 10:15 exam3_metadata.csv
-rw-r--r--  1 jamescutler  staff   813K Apr 24 10:15 exam3_otu_table.csv
STUC02MMPMLFH00:Exam_3 jamescutler$ fastq_to_fasta -i Sample1.fastq -o Sample1.fasta
STUC02MMPMLFH00:Exam_3 jamescutler$ fastq_to_fasta -i Sample2.fastq -o Sample2.fasta
STUC02MMPMLFH00:Exam_3 jamescutler$ ls -ahl
total 31360
drwxr-xr-x  10 jamescutler  staff   320B Apr 26 21:25 .
drwxr-xr-x   5 jamescutler  staff   160B Apr 24 10:15 ..
-rw-r--r--   1 jamescutler  staff   3.4K Apr 24 10:15 Exam_3.Rmd
-rw-r--r--   1 jamescutler  staff   723K Apr 24 10:15 Exam_3.html
-rw-r--r--   1 jamescutler  staff   2.2M Apr 26 21:25 Sample1.fasta
-rw-r--r--   1 jamescutler  staff   4.1M Apr 24 10:15 Sample1.fastq
-rw-r--r--   1 jamescutler  staff   2.6M Apr 26 21:25 Sample2.fasta
-rw-r--r--   1 jamescutler  staff   4.9M Apr 24 10:15 Sample2.fastq
-rw-r--r--   1 jamescutler  staff   3.0K Apr 24 10:15 exam3_metadata.csv
-rw-r--r--   1 jamescutler  staff   813K Apr 24 10:15 exam3_otu_table.csv
STUC02MMPMLFH00:Exam_3 jamescutler$ grep -c "^>" *.fasta
Sample1.fasta:6575
Sample2.fasta:7507
STUC02MMPMLFH00:Exam_3 jamescutler$
```
  
## Now I get the files ready for trimming:
```{r}
setwd("/Users/jamescutler/Desktop/Course_Materials/Data_Course/Exams/Exam_3/")
filt.files = dir(path = getwd(), full.names = FALSE, pattern = ".fastq")
filt.files

for (i in filt.files){
  fastqFilter(fn = i, fout = paste0("/Users/jamescutler/Desktop/Course_Materials/Data_Course/Exams/Exam_3/", i, ".trim"), truncLen = 100)
}
```

  
  
### Part II
```{r}
otu = read.csv("/Users/jamescutler/Desktop/Course_Materials/Data_Course/Exams/Exam_3/exam3_otu_table.csv")
meta = read.csv("/Users/jamescutler/Desktop/Course_Materials/Data_Course/Exams/Exam_3/exam3_metadata.csv")
```

## In order to subset them:
```{r}
meta[which(meta$SampleType == "Soil"),1, drop = FALSE] # A nice view of just the (8) soil samples
meta[which(meta$SampleType == "rhizosphere"),1, drop = FALSE] # A nice view of just the (8) rhizo samples
```

## To save them in a vector:
```{r}
soil.vec = meta[which(meta$SampleType == "Soil"),1]; length(soil.vec)
rhizo.vec = meta[which(meta$SampleType == "rhizosphere"),1]; length(rhizo.vec)
```

## To convert the sample names into a usable format for the purposes of what's coming up next:
```{r}
soil.str = as.character(soil.vec); soil.str
rhizo.str = as.character(rhizo.vec); rhizo.str
```

## Now to manipulate the otu data frame:
## Extracting just the soil and rhizo columns from the otu data frame (".s" suffix is for "subset"):
```{r}
otu.s = otu[,c(soil.str,rhizo.str)]
```

## Transposing the otu subset above so as to have the column names become row names (".t" suffix is for "transpose"):
```{r}
otu.s.t = as.data.frame(t(otu.s))
```

## Now to subset the meta data drame:
```{r}
meta.s = meta[c(which(meta$SampleType == "Soil"),which(meta$SampleType == "rhizosphere")),]
```

## To verify that both subsetted dataframes have the same row names / SampleID names:
```{r}
identical(rownames(otu.s.t), as.character(meta.s$SampleID))
```

## Convert the sample IDs in the meta df to row names:
```{r}
meta.s$SampleID = as.character(meta.s$SampleID)
identical(rownames(otu.s.t), meta.s$SampleID) 
rownames(meta.s) = meta.s$SampleID
identical(rownames(otu.s.t), rownames(meta.s)) # Double check the row names are indeed the same
```

## PermANOVA:
```{r}
pnova1 = adonis(otu.s.t ~ meta.s$SampleType)
```

## I'll throw in a REAL table for ya:
```{r}
adonis.table = data.frame(matrix(NA, nrow = 3, ncol = 6))
for (i in 1:nrow(adonis.table)){
  for (j in 1:ncol(adonis.table)){
    adonis.table[i,j] = pnova1[[1]][i,j]
  }
}  
colnames(adonis.table) = c("Df","SumsOfSqs","MeanSqs","F.Model","R2","Pr(>F)")
rownames(adonis.table) = c("meta.s$SampleType","Residuals","Total")
library(knitr)
kable(adonis.table)
```

# As the PermANOVA table shows, the probability of getting these results under the assumption that sample type makes no difference is .001, so we can probably reject that assumption and conclude that there is sufficient evidence (at a significance level of, say, .05, or even .01) that sample type does make a difference.


### Part III

## Which soil sample is most abundant, and which rhizo sample is most abundant:
## First, soil:
```{r}
otu.soil = otu[,c(soil.str, "Consensus.lineage")]
soil.long = gather(otu.soil, key = "sampleID", value = "OTU_abundance", soil.str)
soil.sums = tapply(soil.long$OTU_abundance, soil.long$sampleID, sum)
soil.sums # Shows 8 abundances. Or.Ma.D.s has the highest abundance.
soil.sums["Or.Ma.D.s"] # Abundance of 34,839
identical(as.numeric(soil.sums["Or.Ma.D.s"]), as.numeric(max(soil.sums))) # confirmed - Or.Ma.D.s is the highest
```

## Next, rhizosphere:
```{r}
otu.rhizo = otu[,c(rhizo.str, "Consensus.lineage")]
rhizo.long = gather(otu.rhizo, key = "sampleID", value = "OTU_abundance", rhizo.str)
rhizo.sums = tapply(rhizo.long$OTU_abundance, rhizo.long$sampleID, sum)
rhizo.sums # Of the 8 abundances shown, the highest is Mg.Sf.D.rz, at 72,337.
rhizo.sums["Mg.Sf.D.rz"]
identical(as.numeric(rhizo.sums["Mg.Sf.D.rz"]), as.numeric(max(rhizo.sums))) # confirmed - Mg.Sf.D.rz is the higest
```
